---
title: "Malaria Infant Paper Analysis"
author: "Paul L. Maurizio"
date: "5/15/2019"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r, message=FALSE}
library(ggplot2)
library(ggfortify)
library(GGally)
library(coin)
library(xtable)
library(pscl) # zero inflated poisson regression
library(MASS) # multinomial ordinal probit regression / proportional odds logistic regression / Box-Cox
library(ordinal) # cumulative link mixed model
library(nparLD) # nonparametric rank-based statistics for longitudinal data
library(lmerTest)
library(psych)
library(coin)
library(cowplot)
```

## Load Data

```{r}
dat <- read.csv("../data/MIS_master_data_sheet_wide.csv")
dat_long <- read.csv("../data/MIS_master_data_sheet_long.csv")
```

## Clinical characteristics

```{r}
dat_children <- subset(dat, age=="I")
dat_adults <- subset(dat, age=="A")
median(dat_children$Hb, na.rm = TRUE); IQR(dat_children$Hb, na.rm = TRUE)
median(dat_adults$Hb, na.rm = TRUE); IQR(dat_adults$Hb, na.rm = TRUE)
median(dat_children$parasites, na.rm = TRUE); IQR(dat_children$parasites, na.rm = TRUE)
median(dat_adults$parasites, na.rm = TRUE); IQR(dat_adults$parasites, na.rm = TRUE)

median(dat_children$pfs25, na.rm = TRUE); IQR(dat_children$pfs25, na.rm = TRUE)
median(dat_adults$pfs25, na.rm = TRUE); IQR(dat_adults$pfs25, na.rm = TRUE)

median(dat_children$pfs16, na.rm = TRUE); IQR(dat_children$pfs16, na.rm = TRUE)
median(dat_adults$pfs16, na.rm = TRUE); IQR(dat_adults$pfs16, na.rm = TRUE)

median(dat_children$pfs230, na.rm = TRUE); IQR(dat_children$pfs230, na.rm = TRUE)
median(dat_adults$pfs230, na.rm = TRUE); IQR(dat_adults$pfs230, na.rm = TRUE)

median(dat_children$Hb.V2, na.rm = TRUE); IQR(dat_children$Hb.V2, na.rm = TRUE)
median(dat_adults$Hb.V2, na.rm = TRUE); IQR(dat_adults$Hb.V2, na.rm = TRUE)
median(dat_children$parasites.V2, na.rm = TRUE); IQR(dat_children$parasites.V2, na.rm = TRUE)
median(dat_adults$parasites.V2, na.rm = TRUE); IQR(dat_adults$parasites.V2, na.rm = TRUE)

```

## PCA on raw phenotypes

This is a PCA on a subset of the phenotypes that were collected from the majority of individuals.

```{r}
select <- c('GMCSF','IFNg','IL10','IL12p40','IL12p70','IL6','TNFa','Nitrate.570','Hb')
dat_sub <- dat_long[,select]
dat_ref <- dat_long[complete.cases(dat_sub),]
dat_sub <- dat_sub[complete.cases(dat_sub),]
dat_ref$Visit <- as.factor(dat_ref$Visit)
dat_ref$AgeVisit <- as.factor(paste(dat_ref$Age, dat_ref$Visit, sep="_"))
pr <- prcomp(dat_sub, center=TRUE, scale.=TRUE)
summary(pr)
pct.variance.explained <- as.numeric(data.frame(summary(pr)$importance)[2,])
x <- c(1:9)
plot(pct.variance.explained ~ x, xlab="PC", ylab="proportion variance explained", las=1, type="b")

autoplot(pr, data = dat_ref, colour = 'AgeVisit', x=1, y=2, loadings=TRUE, loadings.label=TRUE)
autoplot(pr, data = dat_ref, colour = 'AgeVisit', x=2, y=3, loadings=TRUE, loadings.label=TRUE)
```

## Parasite Load

### Model the effects of age, considering sex, on parasite levels on V1, using the Wilcoxon test, among detectable samples only.

```{r}
dat.sub <- subset(dat, parasites>0)
fit1 <- wilcox_test(dat.sub$parasites ~ dat.sub$age | dat.sub$sex, distribution="exact"); pvalue(fit1)
```

### Model the effects of age and sex on parasite levels on V1, using zero-inflated Poisson regression.

We use the zero-inflated Poisson (ZIP) model (log link), with the binomial distribution to model the binary outcome of 0-inflation or not (probit link) (Zeileis 2008).

```{r}
fit1 <- zeroinfl(round(parasites) ~ age * sex , data = dat, dist="poisson", link="probit")
summary(fit1)
```

### Model the effects of age on parasite levels on V2, using zero-inflated Poisson regression.

```{r}
fit1 <- zeroinfl(round(parasites.V2) ~ age, data = dat, dist="poisson", link="probit")
summary(fit1)
```

## Gametocytes

Check the correlation of the Pfs phenotypes.

```{r}
p <- corr.test(log(dat[,c("pfs16", "pfs25", "pfs230")]))$p; p
```

### Model the effects of age and sex on pfs16, pfs25, and pfs230 levels on V1 using the Wilcoxon test

```{r}
fit1 <- wilcox_test(dat$pfs16 ~ dat$age | dat$sex, distribution="exact"); pvalue(fit1)
fit2 <- wilcox_test(dat$pfs25 ~ dat$age | dat$sex, distribution="exact"); pvalue(fit2)
fit3 <- wilcox_test(dat$pfs230 ~ dat$age | dat$sex, distribution="exact"); pvalue(fit3)
median(subset(dat,age=="A")$pfs25, na.rm=TRUE); median(subset(dat,age=="I")$pfs25, na.rm=TRUE);
```

### Check whether the results are similar under distributional assumptions, using lm()

```{r}
fit1 <- lm(log(dat$pfs16) ~ age*sex, data=dat); anova(fit1)
fit2 <- lm(log(dat$pfs25) ~ age*sex, data=dat); anova(fit2)
fit3 <- lm(log(dat$pfs230) ~ age*sex, data=dat); anova(fit3)
```

## Antimalarial antibody

### Model the effects of age and sex on antibody test results at V1 using multinomial ordinal logistic regression

```{r}
dat.sub <- subset(dat, !is.na(malaria.Ab.result))
levels(dat.sub$malaria.Ab.result) <- c("neg", "grey", "pos")
with(dat.sub, table(malaria.Ab.result, sex, age))
fit1 <- polr(malaria.Ab.result ~ age*sex, data=dat.sub, method="logistic")
fit2 <- polr(malaria.Ab.result ~ age + sex, data=dat.sub, method="logistic")
fit3 <- polr(malaria.Ab.result ~ sex, data=dat.sub, method="logistic")
fit4 <- polr(malaria.Ab.result ~ 1, data=dat.sub, method="logistic")
anova(fit4,fit3,fit2,fit1)
```

### Model the effects of age and sex on antibody test results at V2 using multinomial logistic regression.

```{r}
dat.sub <- subset(dat, !is.na(malaria.Ab.result.V2))
levels(dat.sub$malaria.Ab.result.V2) <- c("neg", "grey", "pos")
with(dat.sub, table(malaria.Ab.result.V2, sex, age))
fit1 <- polr(malaria.Ab.result.V2 ~ age*sex, data=dat.sub, method="logistic")
fit2 <- polr(malaria.Ab.result.V2 ~ age + sex, data=dat.sub, method="logistic")
fit3 <- polr(malaria.Ab.result.V2 ~ sex, data=dat.sub, method="logistic")
fit4 <- polr(malaria.Ab.result.V2 ~ 1, data=dat.sub, method="logistic")
anova(fit4,fit3,fit2,fit1)
```

### Model the effects of age, sex, and visit, together, using a cumulative link mixed model (CLMM)/ordinal probit regression.

```{r}
dat.sub <- subset(dat_long, !is.na(malaria.Ab.result))
levels(dat.sub$malaria.Ab.result) <- c("neg", "grey", "pos")
with(dat.sub, table(malaria.Ab.result, Sex, Age, Visit))

fit1 <- clmm(malaria.Ab.result ~ Age*Sex*Visit + (1|Subject_ID), data=dat.sub, Hess=TRUE, link="probit",
              nAGQ=10)
summary(fit1)
```

## Plasma cytokines

### Model the effects of age, sex, and visit on plasma cytokine levels using nparLD (nonparametric).

```{r, results="hide"}
data <- dat_long
host.secreted <- c('TNFa', 'IFNg', 'IL6', 'IL12p40', 'IL12p70', 'IL10',
                   'GMCSF', 'Hb', 'Nitrate.570')

fits <- NULL
fit.df <- as.data.frame(matrix(data=NA, nrow=length(host.secreted),ncol=7))
colnames(fit.df) <- c("Age", "Sex", "Visit", "Age:Sex", "Age:Visit", "Sex:Visit", "Age:Sex:Visit")
fit.df.wholeplot <- as.data.frame(matrix(data=NA, nrow=length(host.secreted),ncol=3))
colnames(fit.df.wholeplot) <- c("Age", "Sex", "Age:Sex")

for(i in 1:length(host.secreted)){
  phen <- host.secreted[i]
  tempdata <- droplevels(subset(data, !is.na(data[phen])))
  complete.subjects <- names(summary(data$Subject_ID))[summary(data$Subject_ID)==2]
  tempdata <- droplevels(subset(data, Subject_ID %in% complete.subjects))
  y <- tempdata[,phen]
  
  time <- tempdata$Visit
  group1 <- tempdata$Age
  group2 <- tempdata$Sex
  subject <- tempdata$Subject_ID
  time.name <- "Visit"
  group1.name <- "Age"
  group2.name <- "Sex"
  fit <-  f2.ld.f1(y=y, time=time, group1=group1, group2=group2,
                   subject=subject, time.name=time.name, 
                   group1.name=group1.name,
                   group2.name=group2.name, plot.RTE=FALSE)
  fits[[i]] <- fit
  fit.df[i,] <- fit$ANOVA.test$`p-value`
  fit.df.wholeplot[i,] <- fit$ANOVA.test.mod.Box$`p-value`
}

cat(paste0(phen, ": \n")); #print(fit$Wald.test); 
rownames(fit.df) <- rownames(fit.df.wholeplot) <- host.secreted
print(fit.df)
xtable(format(fit.df, scientific = TRUE, digits=4))
print(fit.df.wholeplot)
```

### Plot the p-values, colored by significance thresholds

```{r}
par(oma=c(0.1,3,0.1,0.1), mar=c(5,3,0.5,0.5))
plot(1~1, xlim=c(0.5,7.5), ylim=rev(c(1,length(host.secreted)+1)), col="white", ylab="", xlab="",axes=FALSE)
for(i in 1:7){
  for(j in 1:length(host.secreted)){
    if(fit.df[j,i] < 0.05){
      points(x=i,y=j, col=adjustcolor("darkblue", alpha.f=0.15), pch=15, cex=3)
    }
    if(fit.df[j,i] < 0.01){
      points(x=i,y=j, col=adjustcolor("darkblue", alpha.f=0.45), pch=15, cex=3)
    }
    if(fit.df[j,i] < 0.001){
      points(x=i,y=j, col="darkblue", pch=15, cex=3)
    }
  }
}
new.rownames <- c(expression(paste("TNF-",alpha)), expression(paste("IFN-", gamma)), "IL-6", "IL-12 p40", 
                  "IL-12 p70", "IL-10", "GM-CSF", "hemoglobin", "nitric oxide")
axis(side=2,at=c(1:length(host.secreted)),labels=new.rownames, las=2)
text(seq(1, 7, by=1)-0.1, par("usr")[3] + 0.5, labels = colnames(fit.df), 
     srt = 45, pos = 1, offset=1.5, xpd = TRUE)
axis(side=1, at=c(1:7), labels=rep("",7), las=1)
```

## Cell composition phenotypes

### Model the effects of age, sex, and visit on cellular phenotypes using nparLD (nonparametric).

```{r, results="hide"}
data <- dat_long
host.cellular <- c('CD33.live', 'mDC.live', 'monocytes.live', 'inflam.CD163', 
                   'patrol.CD163', 'trad.CD163', 'low.traditional')
fits <- NULL
fit.df <- as.data.frame(matrix(data=NA, nrow=length(host.cellular),ncol=7))
colnames(fit.df) <- c("Age", "Sex", "Visit", "Age:Sex", "Age:Visit", "Sex:Visit", "Age:Sex:Visit")
fit.df.wholeplot <- as.data.frame(matrix(data=NA, nrow=length(host.cellular),ncol=3))
colnames(fit.df.wholeplot) <- c("Age", "Sex", "Age:Sex")

for(i in 1:length(host.cellular)){
  phen <- host.cellular[i]
  tempdata <- droplevels(subset(data, !is.na(data[phen])))
  complete.subjects <- names(summary(data$Subject_ID))[summary(data$Subject_ID)==2]
  tempdata <- droplevels(subset(data, Subject_ID %in% complete.subjects))
  y <- tempdata[,phen]
  
  time <- tempdata$Visit
  group1 <- tempdata$Age
  group2 <- tempdata$Sex
  subject <- tempdata$Subject_ID
  time.name <- "Visit"
  group1.name <- "Age"
  group2.name <- "Sex"
  fit <-  f2.ld.f1(y=y, time=time, group1=group1, group2=group2,
                   subject=subject, time.name=time.name, 
                   group1.name=group1.name,
                   group2.name=group2.name, plot.RTE=FALSE)
  fits[[i]] <- fit
  fit.df[i,] <- fit$ANOVA.test$`p-value`
  fit.df.wholeplot[i,] <- fit$ANOVA.test.mod.Box$`p-value`
}

cat(paste0(phen, ": \n")); #print(fit$Wald.test); 
rownames(fit.df) <- rownames(fit.df.wholeplot) <- host.cellular
```

```{r}
xtable(format(fit.df, scientific = TRUE, digits=4))
print(fit.df.wholeplot)
```

### Plot the p-values (colored by significance thresholds)

```{r}
par(oma=c(0.1,0.1,0.1,0.1), mar=c(5,13,0.5,0.5))
plot(1~1, xlim=c(0.5,7.5), ylim=rev(c(1,8)), col="white", ylab="", xlab="",axes=FALSE)
for(i in 1:7){
  for(j in 1:7){
    if(fit.df[j,i] < 0.05){
      points(x=i,y=j, col=adjustcolor("darkblue", alpha.f=0.15), pch=15, cex=3)
    }
    if(fit.df[j,i] < 0.01){
      points(x=i,y=j, col=adjustcolor("darkblue", alpha.f=0.45), pch=15, cex=3)
    }
    if(fit.df[j,i] < 0.001){
      points(x=i,y=j, col="darkblue", pch=15, cex=3)
    }
  }
}
new.rownames <- c("CD33+ (% of live)",
                  "mDC (% of live)",
                  "monocytes (% of live)",
                  "infl. mono. (% of CD163+)",
                  "patrol. mono. (% of CD163+)",
                  "trad. mono. (% of CD163+)",
                  "CD14^low (% of trad. mono.)")
axis(side=2,at=c(1:7),labels=new.rownames, las=2)
text(seq(1, 7, by=1)-0.1, par("usr")[3], labels = colnames(fit.df), 
     srt = 45, pos = 1, offset=1.5, xpd = TRUE)
axis(side=1, at=c(1:7), labels=rep("",7), las=1)
```

In order to use a parametric (linear mixed model) with our data (lmer) we need to deal with heteroskedastic residuals. We can find a power transform that helps normalize them using Box-Cox analysis (Box and Cox, 1964).

```{r, echo=FALSE}
bcphens <- c("TNFa", "IFNg", "IL6", "IL12p40", "IL12p70",
             "IL10", "GMCSF", "Hb", "Nitrate.570")

boxcoxlist <- list()

par(mfrow=c(3,3))
for(i in 1:length(bcphens)){
  thisdat <- subset(dat_long, select=c("Age", "Sex", "Visit", bcphens[i]))
  thisdat <- thisdat[complete.cases(thisdat),]
  thisdat$Visit <- as.factor(thisdat$Visit)
  thismodel <- paste0(bcphens[i], " ~ Visit*Sex*Age")
  fit1 <- lm(thismodel, data=thisdat, na.action=na.omit)
  boxcoxlist[[i]] <- boxcox(fit1, data = thisdat, lambda = seq(-1,2,0.01), plotit = TRUE, 
                            xlab=bcphens[i]); abline(v=0, col="red"); abline(v=1, col="green")
}

par(mfrow=c(3,3))
for(i in 1:length(bcphens)){
  thisdat <- subset(dat_long, select=c("Age", "Sex", "Visit", bcphens[i]))
  thisdat <- thisdat[complete.cases(thisdat),]
  thisdat$Visit <- as.factor(thisdat$Visit)
  thismodel <- paste0(bcphens[i], " ~ Visit*Sex*Age")
  fit1 <- lm(thismodel, data=thisdat, na.action=na.omit)
  try(plot(fit1, which=2, main=bcphens[i]))
}

```

### Based on the Box-Cox analysis, choose a sensible transform proximal to the lambda value (+/- sqrt, +/- 1/3 root, log, no transform, etc.): Use log (natural) for lambda ~ 0, and no transform for lambda ~ 1.

```{r, echo=FALSE}
bc.transform <- function(y, lambda){(y^lambda - 1)/lambda}

dat_long$GMCSF.bc <- bc.transform(dat_long$GMCSF, lambda=-1/3)
dat_long$IFNg.bc <- bc.transform(dat_long$IFNg, lambda=-0.5)
dat_long$IL10.bc <- bc.transform(dat_long$IL10, lambda=-1/3)
dat_long$IL12p40.bc <- bc.transform(dat_long$IL12p40, lambda=-1/3)
dat_long$IL12p70.bc <- log(dat_long$IL12p70)
dat_long$IL6.bc <- bc.transform(dat_long$IL6, lambda=-0.5)
dat_long$TNFa.bc <- log(dat_long$TNFa)
dat_long$Hb.bc <- dat_long$Hb
dat_long$Nitrate.570.bc <- log(dat_long$Nitrate.570)

bcphens.t <- paste(bcphens, "bc", sep=".")
par(mfrow=c(3,3))
for(i in 1:length(bcphens.t)){
  thisdat <- subset(dat_long, select=c("Age", "Sex", "Visit", bcphens.t[i]))
  thisdat <- thisdat[complete.cases(thisdat),]
  thisdat$Visit <- as.factor(thisdat$Visit)
  thismodel <- paste0(bcphens.t[i], " ~ Visit*Sex*Age")
  fit1 <- lm(thismodel, data=thisdat, na.action=na.omit)
  boxcoxlist[[i]] <- boxcox(fit1, data = thisdat, lambda = seq(-1,2,0.01), plotit = TRUE, 
                            xlab=bcphens.t[i]); abline(v=0, col="red"); abline(v=1, col="green")
}

bcphens.t <- paste(bcphens, "bc", sep=".")
par(mfrow=c(3,3))
for(i in 1:length(bcphens.t)){
  thisdat <- subset(dat_long, select=c("Age", "Sex", "Visit", bcphens.t[i]))
  thisdat <- thisdat[complete.cases(thisdat),]
  thisdat$Visit <- as.factor(thisdat$Visit)
  thismodel <- paste0(bcphens.t[i], " ~ Visit*Sex*Age")
  fit1 <- lm(thismodel, data=thisdat, na.action=na.omit)
  try(plot(fit1, which=2, main=bcphens.t[i]))
}
```

### Use `lmer`:

```{r, results="hide", message=FALSE, warning=FALSE}
dat_long$Visit <- dat_long$Visit - 1 
fits1 <- fits2 <- summaries <- list()
row_names <- c("(Intercept)", "AgeI", "SexM", "Visit", "AgeI:SexM",
               "AgeI:Visit", "SexM:Visit", "AgeI:SexM:Visit")
summary_table <- data.frame(row.names = row_names)
#for(i in c(1:10,18:20)){
for(i in 1:length(bcphens.t)){
  # Subject-level random intercepts will absorb all the age-specific variation, so we leave them out
  # and instead estimate global age-specific effects, and only model the within-subject (visit) slopes
  expr1 <- paste0(bcphens.t[i] , "~ Sex*Visit + (0+Visit|Subject_ID)")
  expr2 <- paste0(bcphens.t[i], "~ Age*Sex*Visit + (0+Visit|Subject_ID)")
  fit1 <- lmer(expr1, data=dat_long, na.action=na.exclude)
  fit2 <- lmer(expr2, data=dat_long, na.action=na.exclude)
  fits1[[i]] <- fit1
  names(fits1)[i] <- bcphens.t[i]
  fits2[[i]] <- fit2
  names(fits2)[i] <- bcphens.t[i]
  cat("\n##-------------------------")
  cat(paste0(as.character(bcphens.t[i]), " : "))
  cat("-------------------------##\n")
  cat("\n##------------")
  cat("SUMMARY")
  cat("--------------##\n")
  print(summary(fit2))
  cat("\n##------------")
  cat("ANOVA")
  cat("--------------##\n")
  print(anova(fit2,fit1))
  cat("\n##------------")
  cat("RANOVA")
  cat("--------------##\n")
  print(ranova(fit2))
  summaries[[i]] <- as.data.frame(summary(fit2)[[10]][,5])
  summary_table <- cbind(summary_table, p=summaries[[i]])
}

colnames(summary_table) <- bcphens
summary_table <- t(summary_table)
xtable(format(summary_table, scientific = TRUE, digits=4))
dat_long$Visit <- dat_long$Visit + 1
```

## Cytokine ratios

### Model the effects of age, sex, and visit on blood analyte ratios using nparLD; we omit IL12p40, NO and Hb, resulting in 15 proportions tested.

```{r, results="hide", message=FALSE, warning=FALSE}
ratiotest <- c('TNFa','IFNg','IL6','IL12p70','IL10','GMCSF')
dat.ratios <- dat_long[,c("Subject_ID", "Sample", "age.years", "Age", "Sex", "Visit")]
ratio.combos <- t(combn(ratiotest,2))
ratio.colnames <- paste(ratio.combos[,1], ratio.combos[,2], sep="/")
for(i in 1:length(ratio.colnames)){
  dat.ratios[,ratio.colnames[i]] <- dat_long[,ratio.combos[i,1]]/dat_long[,ratio.combos[i,2]]
}

fits <- NULL
fit.df <- as.data.frame(matrix(data=NA, nrow=length(ratio.colnames),ncol=7))
colnames(fit.df) <- c("Age", "Sex", "Visit", "Age:Sex", "Age:Visit", "Sex:Visit", "Age:Sex:Visit")
fit.df.wholeplot <- as.data.frame(matrix(data=NA, nrow=length(ratio.colnames),ncol=3))
colnames(fit.df.wholeplot) <- c("Age", "Sex", "Age:Sex")

for(i in 1:length(ratio.colnames)){
  phen <- ratio.colnames[i]
  cat(paste0(phen, ": \n")); #print(fit$Wald.test); 
  #tempdata <- droplevels(subset(dat.ratios, !is.na(data[phen])))
  complete.subjects <- names(summary(data$Subject_ID))[summary(data$Subject_ID)==2]
  tempdata <- droplevels(subset(dat.ratios, Subject_ID %in% complete.subjects))
  y <- tempdata[,phen]
  
  time <- tempdata$Visit
  group1 <- tempdata$Age
  group2 <- tempdata$Sex
  subject <- tempdata$Subject_ID
  time.name <- "Visit"
  group1.name <- "Age"
  group2.name <- "Sex"
  fit <-  f2.ld.f1(y=y, time=time, group1=group1, group2=group2,
                   subject=subject, time.name=time.name, 
                   group1.name=group1.name,
                   group2.name=group2.name, plot.RTE=FALSE)
  fits[[i]] <- fit
  fit.df[i,] <- fit$ANOVA.test$`p-value`
  fit.df.wholeplot[i,] <- fit$ANOVA.test.mod.Box$`p-value`
}

rownames(fit.df) <- rownames(fit.df.wholeplot) <- ratio.colnames
```

```{r}
xtable(format(fit.df, scientific = TRUE, digits=4))
print(fit.df)
print(fit.df.wholeplot)
```

### Plot the p-values in a grid.

```{r, echo=FALSE}
par(oma=c(0.1,0.1,0.1,0.1), mar=c(5,9,0.5,0.5))
plot(1~1, xlim=c(0.5,7.5), ylim=rev(c(1,nrow(fit.df))), col="white", ylab="", xlab="",axes=FALSE)
for(i in 1:7){
  for(j in 1:nrow(fit.df)){
    if(fit.df[j,i] < 0.05){
      points(x=i,y=j, col=adjustcolor("darkblue", alpha.f=0.15), pch=15, cex=3)
    }
    if(fit.df[j,i] < 0.01){
      points(x=i,y=j, col=adjustcolor("darkblue", alpha.f=0.45), pch=15, cex=3)
    }
    if(fit.df[j,i] < 0.001){
      points(x=i,y=j, col="darkblue", pch=15, cex=3)
    }
  }
}
axis(side=2,at=c(1:nrow(fit.df)),labels=rownames(fit.df), las=2)
text(seq(1, 7, by=1)-0.1, par("usr")[3], labels = colnames(fit.df), 
     srt = 45, pos = 1, offset=1.5, xpd = TRUE)
axis(side=1, at=c(1:7), labels=rep("",7), las=1)
```

## Treatment failure, non-compliance, or recrudescence.

```{r, echo=FALSE}
dat$color.by.age <- ifelse(dat$age=="A", rgb(0,0,0,1,alpha=0.25), rgb(1,0,0,1,alpha=0.25))
dat$color.by.age1 <- ifelse(dat$age=="A", rgb(0,0,0,1), rgb(1,0,0,1))
dat$shape.by.sex <- ifelse(dat$sex=="M", 24, 21)
dat$shape.by.sex1 <- ifelse(dat$sex=="M", 17, 16)
dat$treatfail <- ifelse(dat$parasites.V2>10,1,0)

phens.sub <- c("parasites", "Hb", "Nitrate.570", "TNFa", "IFNg", "IL6", "IL12p40", "IL12p70",
                   "IL10", "GMCSF")
new.rownames <- c("parasites", "hemoglobin", "nitric oxide", expression(paste("TNF-",alpha)), expression(paste("IFN-", gamma)), "IL-6", "IL-12 p40", "IL-12 p70",
                  "IL-10", "GM-CSF")

par(mfrow=c(4,3), mar=c(3,3,2.5,1))
i <- 1
phen.v1 <- phens.sub[i]
phen.v2 <- paste0(phens.sub[i],".V2")
expr <- substitute(log(j+1,base=10) ~ log(i+1,base=10), list(i = as.name(phen.v1), j=as.name(phen.v2)))
plot(eval(expr), data=dat, main=new.rownames[i], xlab="", ylab="", pch=shape.by.sex1, col=color.by.age, cex=1.5,
     ylim=range(log(dat[,phen.v1]+1, base=10), na.rm=TRUE), las=1)
points(eval(expr), data=subset(dat, treatfail==1), main=new.rownames[i], xlab="", ylab="", pch=shape.by.sex, bg=color.by.age1, col="black", cex=1.5,
       ylim=range(log(dat[,phen.v1]+1, base=10), na.rm=TRUE))
abline(a=0,b=1)
for(i in 2:(length(phens.sub))) {
  phen.v1 <- phens.sub[i]
  phen.v2 <- paste0(phens.sub[i],".V2")
  expr <- substitute(log(j,base=10) ~ log(i,base=10), list(i = as.name(phen.v1), j=as.name(phen.v2)))
  plot(eval(expr), data=dat, main=new.rownames[i], xlab="", ylab="", pch=shape.by.sex1, col=color.by.age, cex=1.5, las=1)
  points(eval(expr), data=subset(dat, treatfail==1), main=new.rownames[i], xlab="", ylab="", pch=shape.by.sex, bg=color.by.age1, col="black", cex=1.5)
  abline(a=0,b=1)
}
```

## Continuous age-associated effects within group

### Model the phenotypes vs. age for adults and young children at V1.

```{r, results="hide"}
phens <- c('GMCSF', 'IFNg', 'IL10', 'IL12p40', 'IL12p70', 
           'IL6', 'TNFa', 'Nitrate.570',  
           'pfs25', 'pfs16', 'pfs230', 'Hb', 'malaria.Ab', 'parasites')

phens <- phens[!(phens=="malaria.Ab")]
dat.I <- subset(dat, age="I")
dat.A <- subset(dat, age="A")

cat("##-----------------##\n")
cat("CHILDREN \n")
cat("##-----------------##\n")
for(i in 1:length(phens)){
  cat("##-----------------")
  cat(phens[i])
  cat("-----------------## \n")
  form <- paste0(phens[i], "~ age.years*sex")
  try(fit1 <- lm(form, data=dat.I))
  try(print(summary(fit1)))
  fit1 <- NULL
}

for(i in 1:length(phens)){
  cat("##-----------------")
  cat(phens[i])
  cat("-----------------## \n")
  form <- paste0(phens[i], "~ age.years*sex")
  try(fit1 <- lm(form, data=dat.A))
  try(print(summary(fit1)))
  fit1 <- NULL
}
```

### Model the phenotypes vs. age for adults and young children at V2.

```{r, results="hide"}
phens.V2 <- paste0(phens, ".V2")
phens.V2 <- phens.V2[!(phens.V2 %in% c("pfs25.V2", "pfs16.V2", "pfs230.V2", "malaria.Ab.V2"))]

cat("##-----------------##\n")
cat("CHILDREN \n")
cat("##-----------------##\n")
for(i in 1:length(phens.V2)){
  cat("##-----------------")
  cat(phens.V2[i])
  cat("-----------------## \n")
  form <- paste0(phens.V2[i], "~ age.years*sex")
  try(fit1 <- lm(form, data=dat.I))
  try(print(summary(fit1)))
  fit1 <- NULL
}

cat("##-----------------##\n")
cat("ADULTS \n")
cat("##-----------------##\n")
for(i in 1:length(phens.V2)){
  cat("##-----------------")
  cat(phens.V2[i])
  cat("-----------------## \n")
  form <- paste0(phens.V2[i], "~ age.years*sex")
  try(fit1 <- lm(form, data=dat.A))
  try(print(summary(fit1)))
  fit1 <- NULL
}
```

### Model the log2FC of phenotypes vs. age for adults and young children.

```{r, results="hide"}
dat$TNFa.FC <- log2(dat$TNFa.V2) - log2(dat$TNFa)
dat$IFNg.FC <- log2(dat$IFNg.V2) - log2(dat$IFNg)
dat$IL6.FC <- log2(dat$IL6.V2) - log2(dat$IL6)
dat$IL12p40.FC <- log2(dat$IL12p40.V2) - log2(dat$IL12p40)
dat$IL12p70.FC <- log2(dat$IL12p70.V2) - log2(dat$IL12p70)
dat$IL10.FC <- log2(dat$IL10.V2) - log2(dat$IL10)
dat$GMCSF.FC <- log2(dat$GMCSF.V2) - log2(dat$GMCSF)
dat$Hb.FC <- log2(dat$Hb.V2) - log2(dat$Hb)
dat$Nitrate.570.FC <- log2(dat$Nitrate.570.V2) - log2(dat$Nitrate.570)
phens <- c("TNFa", "IFNg", "IL6", "IL12p40", "IL12p70", "IL10", "GMCSF", "Hb", "Nitrate.570")
phens.FC <- paste0(phens, ".FC")
dat.I <- subset(dat, age="I")
dat.A <- subset(dat, age="A")

cat("##-----------------##\n")
cat("CHILDREN \n")
cat("##-----------------##\n")
for(i in 1:length(phens.FC)){
  cat("##-----------------")
  cat(phens.FC[i])
  cat("-----------------## \n")
  form <- paste0(phens.FC[i], "~ age.years*sex")
  try(fit1 <- lm(form, data=dat.I))
  try(print(summary(fit1)))
  fit1 <- NULL
}

cat("##-----------------##\n")
cat("ADULTS \n")
cat("##-----------------##\n")
for(i in 1:length(phens.FC)){
  cat("##-----------------")
  cat(phens.FC[i])
  cat("-----------------## \n")
  form <- paste0(phens.FC[i], "~ age.years*sex")
  try(fit1 <- lm(form, data=dat.A))
  try(print(summary(fit1)))
  fit1 <- NULL
}
```

### Plot the phenotypes vs. age for adults and young children - V1.

```{r, message=FALSE, warning=FALSE}
dat.I <- droplevels(subset(dat, age=="I"))
dat.A <- droplevels(subset(dat, age=="A"))

p1 <- ggplot(dat.I, aes_string(x="age.years", y="TNFa", colour="sex", shape="sex"))
p2 <- p1 + geom_point(aes(pch=sex), size=3) + geom_smooth(method=lm) + theme_classic() + geom_hline(yintercept=0, color = "darkgrey") + ylab(expression(paste("TNF-",alpha," (pg/mL)")))
plot(p2)

p1 <- ggplot(dat.A, aes_string(x="age.years", y="pfs16", colour="sex", shape="sex"))
p2 <- p1 + geom_point(aes(pch=sex), size=3) + geom_smooth(method=lm) + ylim(-75,200) + theme_classic() + geom_hline(yintercept=0, color = "darkgrey") + ylab("Pfs16 (gametocytes/uL)")
plot(p2)
```

### Plot the phenotypes vs. age for adults and young children - V2.

```{r, message=FALSE, warning=FALSE}
p1 <- ggplot(dat.A, aes_string(x="age.years", y="Nitrate.570.V2", colour="sex", shape="sex"))
p2 <- p1 + geom_point(aes(pch=sex), size=3) + geom_smooth(method=lm) + theme_classic() + geom_hline(yintercept=0, color = "darkgrey") + ylab("nitric oxide (uM)")
plot(p2)
```

## Comparison of parasites and gameocytes at V1

Compare estimated levels of gametocytemia in samples with parasitemia = 0 and parasitemia > 0.

```{r, echo=FALSE}
dat_long$parasites_detected <- as.factor(ifelse(dat_long$parasites > 0, TRUE, ifelse(dat_long$parasites < 1, FALSE, NA)))

# subset on data for which parasite counts were obtained

dat_sub <- droplevels(subset(dat_long, !is.na(parasites_detected)))
dat_sub$AgeSex <- as.factor(paste(dat_sub$Age, dat_sub$Sex, sep="_"))

p1 <- ggplot(dat_sub, aes(x=parasites_detected, y=pfs16)) + scale_y_log10() +   geom_boxplot(outlier.shape=NA) + theme_classic() + 
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5)
p2 <- ggplot(dat_sub, aes(x=parasites_detected, y=pfs25)) + scale_y_log10() +
  geom_boxplot(outlier.shape=NA) + theme_classic() + 
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5)
p3 <- ggplot(dat_sub, aes(x=parasites_detected, y=pfs230)) + scale_y_log10() + 
  geom_boxplot(outlier.shape=NA) + theme_classic() + 
  geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5)

#pdf("parasitemia_gametocytemia.pdf", width=6, height=4)
plot_grid(p1, p2, p3, ncol = 3, labels=LETTERS[1:3])
#dev.off()

fit1 <- wilcox_test(dat_sub$pfs25 ~ dat_sub$parasites_detected | dat_sub$AgeSex, distribution="exact"); pvalue(fit1)
fit2 <- wilcox_test(dat_sub$pfs16 ~ dat_sub$parasites_detected | dat_sub$AgeSex, distribution="exact"); pvalue(fit2)
fit3 <- wilcox_test(dat_sub$pfs230 ~ dat_sub$parasites_detected | dat_sub$AgeSex, distribution="exact"); pvalue(fit3)
```

## Model the effect of parasite load at V1 on the transformed phenotypes that we examined previously.

```{r, echo=FALSE}
dat_long$scaled.parasites <- scale(dat_long$parasites, center=TRUE, scale=TRUE)

dat_long_V1 <- droplevels(subset(dat_long, Visit==1))

fit <- lm(GMCSF.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(IFNg.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(IL10.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(IL12p40.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(IL12p70.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(IL6.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(TNFa.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(Hb.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

fit <- lm(Nitrate.570.bc ~ Age*Sex*scaled.parasites, data=dat_long_V1); anova(fit)[["Pr(>F)"]][3]

# IFNg, IL10, IL6, TNFa

dat$scaled.parasites <- scale(dat$parasites, scale=TRUE, center=TRUE)
fit <- lm(, data=dat)

# parasite levels do significantly affect IFNg, IL-10, IL-6, and TNFa levels

```
